---
title: \textcolor{myblue}{`r "A4 / LEARN Data Primer"`}
output:
  pdf_document:
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage{graphicx}
  - \usepackage{geometry}
  - \geometry{a4paper,includehead,top=0cm}
  - \usepackage{helvet}  % Helvetica font package
  - \renewcommand{\familydefault}{\sfdefault}  % Set default font to sans-serif
  - \usepackage{sectsty}
  - \usepackage{xcolor}
  - \definecolor{myblue}{HTML}{2C7FA0}
  - \allsectionsfont{\color{myblue}}
urlcolor: blue
---

```{r settings, results = "hide", echo = FALSE, message = FALSE, warning = FALSE} 
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
  warning = FALSE, fig.width = 8, fig.height = 8 / 1.6, out.width = "\\linewidth",
  fig.align = "center", tidy = FALSE, comment = NA, cache = FALSE, crop = TRUE,
  echo = TRUE)

options(stringsAsFactors = FALSE, digits = 2, width = 150)
```

```{=tex}
\pagestyle{fancy} 
\setlength\headheight{62pt}
\fancypagestyle{plain}{}
\fancyhf{}
```

```{r, echo=FALSE, results='asis'}
if (file.exists('../methods/A4_LRN_logo.png')) {
  cat("\\chead{\\makebox[\\headwidth][c]{\\includegraphics[width=\\paperwidth]{../methods/A4_LRN_logo.png}}}")
}
```

```{=tex}
\renewcommand{\headrulewidth}{0pt}
```
# Introduction
This document demonstrates an example on how to explore the various clinical datasets made available using R programming language.

## Load required R packages
\footnotesize
```{r packages, results = "hide", message = FALSE, warning = FALSE} 
library(tidyverse)
library(arsenal)
```
\normalsize

## Load CSV data files into R
\footnotesize
```{r load_data, results = "hide"}
datadic <- read.csv("clinical_datadic.csv")     # Data dictionary
# raw dataset
ptdemog <- read.csv("ptdemog.csv")     # Participant demographics
# derived datasets
subjinfo <- read.csv("SUBJINFO.csv") # Subject info (re-screens and APOE status)
pacc <- read.csv("PACC.csv") # PACC
# imaging datasets
petsuvr <- read.csv("imaging_SUVR_amyloid.csv") # Amyloid PET Quantitative results
petvadata <- read.csv("imaging_PET_VA.csv") # Amyloid PET Eligibility
vmri <- read.csv("imaging_volumetric_mri.csv") # Volumetric MRI
```
\normalsize

## Define functions for extracting meta data from data dictionary
\footnotesize
```{r functions, results = "hide"}
get_levels <- function(x){
  as.numeric(unlist(lapply(strsplit(unlist(strsplit(subset(
    datadic, FIELD_NAME==x)$FIELD_CODE, ';')), '='), function(y) y[1])))
}
get_labels <- function(x){
  unlist(lapply(strsplit(unlist(strsplit(subset(
    datadic, FIELD_NAME==x)$FIELD_CODE, ';')), '='), function(y) y[2]))
}
```
\normalsize

# Prepare data

**Rescreens**

Participants who re-screened may appear in the data twice with different BIDs 
each time. The SUBJINFO derived dataset indicates which participants are re-screens,
and how the re-screen BIDs are mapped to each other. The code below also accounts
for this to set up the removal of duplicate appearances.
\footnotesize
```{r prep_rescreens, results = "hide"}
rescreens <- subjinfo %>%
  filter(!is.na(PREVBID)) %>%
  rename(BID1 = PREVBID, BID2 = BID) %>%
  select(BID1, BID2)
```
\normalsize

**Gather Amyloid PET data**

The Amyloid PET quantitative data (`petsuvr`) is in long format with one row per
region. We use `tidyr::pivot_wider` to transform to wide format with one column
per region.
\footnotesize
```{r prep_pet, results = "hide"}
pet <- petsuvr %>% 
  filter(brain_region != '' & VISCODE == 2) %>%
  pivot_wider(id_cols='BID', names_from=brain_region, values_from=suvr_cer) %>%
  left_join(petvadata, by='BID') %>%
  left_join(rescreens, by=c('BID' = 'BID1')) %>% 
  mutate( # update PET BIDs to second BID if necessary
    BID = case_when(
      !is.na(BID2) ~ BID2,
      TRUE ~ BID)) %>%
  arrange(BID, BID2) %>%
  filter(!duplicated(BID, fromLast = TRUE)) %>%
  select(-BID2)
```
\normalsize

**Label Participant Demographics data**

Raw datasets are presented using coded values. Their translated labels can be found 
using the relevant data dictionary. The code below demonstrates how the labelled 
values can be derived using the functions previously defined at the start of this document.
\footnotesize
```{r prep_ptdemog, results = "hide"}
ptdemog <- ptdemog %>%
  mutate(
    PTGENDER = factor(PTGENDER, 
      levels = get_levels('PTGENDER'),
      labels = get_labels('PTGENDER')),
    PTETHNIC = factor(PTETHNIC, 
      levels = get_levels('PTETHNIC'),
      labels = get_labels('PTETHNIC')),
    PTMARRY = factor(PTMARRY, 
      levels = get_levels('PTMARRY'),
      labels = get_labels('PTMARRY')),
    PTNOTRT = factor(PTNOTRT, 
      levels = get_levels('PTNOTRT'),
      labels = get_labels('PTNOTRT')))
```
\normalsize
Some variables were collected as multi-checkbox selections. The data is 
aggregated to a string of numeric values seperated by a colon (e.g. '2:3:5'). The
code below is an example of how to convert the variable to distinct binomial fields.
\footnotesize
```{r prep_ptrace, results = "hide"}
race.levs <- get_levels('PTRACE')
race.labs <- get_labels('PTRACE')
for (i in 1:length(race.levs)){
  ptdemog[!is.na(ptdemog$PTRACE),  paste('PTRACE:',race.labs[i])] <- 0
  ptdemog[grep(race.levs[i], ptdemog$PTRACE, fixed = TRUE), paste('PTRACE:',race.labs[i])] <- 1
}
```
NOTE: Key participant demographics and baseline characteristics can also be 
found in the SUBJINFO derived dataset. PTDEMOG was used here to demonstrate how 
to label raw data using the data dictionary.
\normalsize

**Baseline PACC**
\footnotesize
```{r prep_pacc, results = "hide"}
pacc_bl <- pacc %>% 
  filter(VISCODE == 6) %>%
  select(BID, PACC.raw, MMSCORE, LDELTOTAL, DIGITTOTAL, FCTOTAL96)
```
\normalsize

**Prepare data table**
\footnotesize
```{r prep_data, results = "hide"}
dd <- subjinfo[,c('BID','APOEGN')] %>%
  left_join(ptdemog, by='BID') %>%
  left_join(pet, by=c('SUBSTUDY','BID')) %>%
  left_join(vmri[is.na(vmri$VISCODE) | vmri$VISCODE==4,]) %>%
  left_join(pacc_bl, by=c('BID')) %>%
  filter(!BID %in% rescreens$BID1) %>% # remove first appearance of rescreens
  rename(`A4 amyloid eligibility` = overall_score) %>%
  rename(
    `Age at screening (yrs)` = PTAGE,
    `APOE genotype` = APOEGN, 
    `Education (yrs)` = PTEDUCAT, 
    `PET SUVr` = Composite_Summary,
    Sex = PTGENDER,
    Ethnicity = PTETHNIC,
    `Marital status` = PTMARRY,
    `Participant retired` = PTNOTRT,
    `Hippocampal Occupancy` = HOC,
    PACC = PACC.raw, 
    MMSE = MMSCORE, 
    `Logical memory delay` = LDELTOTAL, 
    `Digit symbol` = DIGITTOTAL, 
    `FCSRT (2xFree + Cued)` = FCTOTAL96
  ) %>% 
  filter(!SUBSTUDY %in% 'SF') # excluding screen-fails
```
\normalsize

\newpage
# Summarize data

\scriptsize
```{r summarize, results='asis'}
dd<- dd %>% # rename values and fields for footnote setup in summary table
  mutate(SUBSTUDY = ifelse(SUBSTUDY == 'A4', 'A4^a^' , SUBSTUDY)) %>%
  rename(
    `A4 amyloid eligibility^b^` = `A4 amyloid eligibility`,
    `Amyloid PET SUVr^b^` = `PET SUVr`
  )
tableby(SUBSTUDY ~ .,
  data = select(dd, SUBSTUDY, `A4 amyloid eligibility^b^`, `Age at screening (yrs)`, `Education (yrs)`,
    `APOE genotype`, `Amyloid PET SUVr^b^`, `Hippocampal Occupancy`, Sex, Ethnicity,
    `Marital status`, `Participant retired`, 
    PACC, MMSE, `Logical memory delay`, `Digit symbol`, `FCSRT (2xFree + Cued)`), digits = 2) %>%
summary(title = "Baseline characteristics of A4-randomized^a^ and LEARN-enrolled cohorts")
```
\scriptsize
^a^ A4-randomized cohort (n=`r sum(dd$SUBSTUDY=='A4^a^')`) includes other participants in addition to the modified intention-to-treat population (mITT n=1147) reported in the A4 trial (Sperling et al. 2023). The modified intention-to-treat population population that was reported for the A4 trial results include those who received at least one dose of solanezumab or placebo and underwent assessment for the primary end point. Please refer to the Intro-to-A4.pdf file for code to reproduce the baseline characteristics and primary findings of the A4 study. \
^b^ Refer to A4 Amyloid PET Eligibility Methods PDF document (imaging_PET_VA_methods.pdf) regarding these amyloid-related measures, the eligibility determination process and the modifications made to the SUVR algorithm.
\normalsize

\newpage
# References
R Core Team. 2019. R: A Language and Environment for Statistical Computing. Vienna, Austria: R
Foundation for Statistical Computing. https://www.R-project.org.

Sperling, R. A., Donohue, M. C., Raman, R., Rafii, M. S., Johnson, K., Masters, C. L., van Dyck, C. H., Iwatsubo, T., Marshall, G. A., Yaari, R., Mancini, M., Holdridge, K. C., Case, M., Sims, J. R., Aisen, P. S., & A4 Study Team (2023). Trial of Solanezumab in Preclinical Alzheimer's Disease. *The New England journal of medicine, 389*(12), 1096–1107. https://doi.org/10.1056/NEJMoa2305032