---
title: "Multivariate Power Analysis with AIC-Optimized Model - Week 48"
author: "Ian Sequeira"
format: html
editor: visual
---

# Load Libraries and Helper Functions

```{r setup}
#| message: false
#| warning: false

library(here)
source(here("helpers", "01-load-libraries.R"))
source(here("helpers", "02-functions.R"))

# Additional libraries specific to this analysis
library(purrr)
library(ggplot2)
library(scales)
library(ggrepel)
library(broom)
library(kableExtra)
library(MASS)  
library(boot)
library(pROC)
```

# Custom Functions

```{r custom_functions}
#| message: false
#| warning: false

# Function to format p-values
formatp <- function(x) case_when(
  x < 0.001 ~ "p<0.001",
  x > 0.01 ~ Hmisc::format.pval(x, digits=2, eps=0.01, nsmall=2),
  TRUE ~ Hmisc::format.pval(x, digits=3, eps=0.001, nsmall=3))

# ROC calculation function for AUC
calculate_roc_auc <- function(y_true, risk_scores) {
  complete_cases <- !is.na(y_true) & !is.na(risk_scores)
  y_true <- y_true[complete_cases]
  risk_scores <- risk_scores[complete_cases]
  
  if(length(unique(y_true)) < 2) return(NA)
  
  cutpt <- seq(min(risk_scores), max(risk_scores), length = 100)
  sensitivity <- specificity <- rep(NA, length(cutpt))
  
  for(i in 1:length(cutpt)) {
    prog <- risk_scores[y_true == 1]
    noprog <- risk_scores[y_true == 0]
    
    sensitivity[i] <- sum(prog >= cutpt[i]) / length(prog)
    specificity[i] <- sum(noprog < cutpt[i]) / length(noprog)
  }
  
  auc <- sum(diff(1 - specificity) * (sensitivity[-1] + sensitivity[-length(sensitivity)]) / 2)
  return(abs(auc))
}

# Power analysis function
calculate_sample_size <- function(variance, delta, alpha = 0.05, beta = 0.20) {
  if(is.na(variance) || is.na(delta) || variance <= 0 || delta == 0) {
    return(NA)
  }
  
  z_alpha <- qnorm(1 - alpha / 2)
  z_beta <- qnorm(1 - beta)
  n <- (2 * variance * (z_alpha + z_beta)^2) / delta^2
  
  return(ceiling(n))
}

# Function to calculate test-specific risk scores
calculate_test_risk_scores <- function(data, test_name, model) {
  test_data <- data
  
  # Identify variables for this test
  if(test_name == "C3") {
    test_vars <- "z_delta_C3_W48"
  } else {
    test_vars <- c(paste0("adjusted_delta_", test_name, "_W48"),
                   paste0("baseline_", test_name))
  }
  
  # Get predictors that are actually in the optimal model
  model_terms <- names(coef(model))[-1]
  all_predictors <- unique(unlist(strsplit(model_terms, ":")))
  
  # Set other predictors to their means
  other_predictors <- setdiff(all_predictors, test_vars)
  
  for(pred in other_predictors) {
    if(pred %in% names(test_data)) {
      test_data[[pred]] <- mean(data[[pred]], na.rm = TRUE)
    }
  }
  
  # Calculate risk scores (log odds)
  risk_scores <- predict(model, newdata = test_data, type = "link")
  return(risk_scores)
}
```

# Data Preparation and Quality Checks

```{r load_data}
#| message: false
#| warning: false

# Load the starting data files
data_wk48 <- read_csv("data/data_wk48.csv")
learn_data <- read_csv("data/data_learn&base_wk48.csv")

# Check data structure
cat("A4 data dimensions:", nrow(data_wk48), "x", ncol(data_wk48), "\n")
cat("LEARN data dimensions:", nrow(learn_data), "x", ncol(learn_data), "\n")
cat("Number of positive outcomes in A4:", sum(data_wk48$CDPOS_W240), "\n")
cat("Outcome rate in A4:", round(mean(data_wk48$CDPOS_W240), 3), "\n\n")

# Define test codes
test_codes <- c("ADLPQPT", "ADLPQSP", "CFIPT", "CFISP", 
                "DIGIT", "FCSRT96", "LMIIA", "MMSE")
```

# LEARN Effect Calculation and Adjustment

```{r learn_effects}
# Calculate LEARN effects (population-level practice effects)
learn_effects <- learn_data %>%
  summarise(
    across(starts_with("z_delta_") & ends_with("_W48"), 
           ~mean(.x, na.rm = TRUE), 
           .names = "learn_effect_{.col}")
  ) %>%
  rename_all(~str_replace(., "z_delta_", ""))

cat("LEARN effects calculated for adjustment:\n")
print(learn_effects)
```

```{r create_adjusted_variables}
# Create adjusted delta scores by subtracting LEARN effects
data_wk48_adjusted <- data_wk48

# For each test, create adjusted delta
for(test in test_codes) {
  z_delta_col <- paste0("z_delta_", test, "_W48")
  learn_effect_col <- paste0("learn_effect_", test, "_W48")
  adjusted_col <- paste0("adjusted_delta_", test, "_W48")
  
  if(z_delta_col %in% names(data_wk48) && learn_effect_col %in% names(learn_effects)) {
    data_wk48_adjusted[[adjusted_col]] <- 
      data_wk48[[z_delta_col]] - learn_effects[[learn_effect_col]]
    cat("Created", adjusted_col, "\n")
  }
}

# Create baseline scores (Week 48 raw scores)
for(test in test_codes) {
  raw_col <- paste0(test, "_W48")
  baseline_col <- paste0("baseline_", test)
  
  if(raw_col %in% names(data_wk48)) {
    data_wk48_adjusted[[baseline_col]] <- data_wk48[[raw_col]]
    cat("Created", baseline_col, "\n")
  }
}

# C3 doesn't need adjustment (not in LEARN)
if("z_delta_C3_W48" %in% names(data_wk48_adjusted)) {
  cat("\nC3 delta score included (no LEARN adjustment needed)\n")
}

# Remove rows with missing outcomes
data_complete <- data_wk48_adjusted %>% 
  filter(!is.na(CDPOS_W240))

cat("\nFinal data dimensions:", nrow(data_complete), "x", ncol(data_complete), "\n")
```

# Individual Test Risk Score Analysis

```{r individual_test_analysis}
#| message: false
#| warning: false

# Calculate results for each test using BIC-optimal model
test_codes_all <- c(test_codes, "C3")
auc_results <- data.frame()

cat("\nCalculating test-specific AUCs using BIC-optimal model...\n")
cat("Model has", length(coef(glm_optimal)), "parameters (including intercept)\n\n")

for(test in test_codes_all) {
  # Check if this test's variables are in the optimal model
  if(test == "C3") {
    test_var <- "z_delta_C3_W48"
  } else {
    test_var <- paste0("adjusted_delta_", test, "_W48")
  }
  
  # Check if test variable is in the model (main effect or interaction)
  model_vars <- names(coef(glm_optimal))
  if(!any(grepl(test_var, model_vars))) {
    cat("\n", test, "not included in BIC-optimal model - skipping\n")
    next
  }
  
  cat("\nProcessing", test, "...\n")
  
  # Get variable names
  if(test == "C3") {
    delta_var <- "z_delta_C3_W48"
    adj_delta_var <- delta_var
  } else {
    delta_var <- paste0("z_delta_", test, "_W48")
    adj_delta_var <- paste0("adjusted_delta_", test, "_W48")
  }
  
  # Skip if variables don't exist in data
  var_col <- ifelse(test == "C3", delta_var, adj_delta_var)
  if(!var_col %in% names(data_complete)) {
    cat("  Variable not found in data - skipping\n")
    next
  }
  
  # Calculate risk scores
  risk_scores <- calculate_test_risk_scores(data_complete, test, glm_optimal)
  
  # Calculate AUC
  auc <- calculate_roc_auc(data_complete$CDPOS_W240, risk_scores)
  
  # Calculate variance of the LINEAR PREDICTOR
  risk_score_variance <- var(risk_scores, na.rm = TRUE)
  
  # Calculate mean change in risk score between positive and negative outcomes
  mean_risk_pos <- mean(risk_scores[data_complete$CDPOS_W240 == 1], na.rm = TRUE)
  mean_risk_neg <- mean(risk_scores[data_complete$CDPOS_W240 == 0], na.rm = TRUE)
  mean_risk_delta <- mean_risk_pos - mean_risk_neg
  
  # Get sample size for reporting
  n_complete <- sum(!is.na(data_complete[[var_col]]))
  
  cat("  AUC:", round(auc, 3), "\n")
  cat("  Risk score variance:", round(risk_score_variance, 4), "\n")
  cat("  Mean risk score difference:", round(mean_risk_delta, 4), "\n")
  cat("  (Original data variance:", round(var(data_complete[[var_col]], na.rm = TRUE), 4), ")\n")
  
  # Add to results - using risk score statistics
  auc_results <- rbind(auc_results, 
                       data.frame(
                         test_name = test,
                         auc = auc,
                         variance = risk_score_variance,
                         mean_delta = mean_risk_delta,
                         n_complete = n_complete,
                         stringsAsFactors = FALSE
                       ))
}

# Sort by AUC
auc_results <- auc_results %>% 
  filter(!is.na(auc) & !is.na(variance) & variance > 0) %>%
  arrange(desc(auc))

cat("\n=== AUC RESULTS SUMMARY ===\n")
print(auc_results)
```

# Power Analysis: Original Effect Sizes

```{r power_analysis_original}
# Calculate sample sizes using original observed effect sizes
power_results <- auc_results %>%
  mutate(
    abs_mean_delta = abs(mean_delta),
    sample_size = map2_dbl(variance, abs_mean_delta, 
                           ~calculate_sample_size(.x, .y, alpha = 0.05, beta = 0.20)),
    test_label = paste0(test_name, "\n(AUC: ", sprintf("%.2f", auc), ")")
  ) %>%
  filter(!is.na(sample_size) & sample_size > 0 & sample_size < 1e6) %>%
  arrange(sample_size)

cat("\nPower Analysis Results (BIC-Optimized Model):\n")
cat("================================================\n")
print(power_results %>% 
      dplyr::select(test_name, auc, sample_size, variance, abs_mean_delta) %>%
      mutate(across(where(is.numeric), ~round(., 3))))

# Check if we have reasonable sample sizes
if(any(power_results$sample_size > 10000)) {
  cat("\nWARNING: Some sample sizes are very large (>10,000).\n")
  cat("This may indicate the BIC model removed too many predictors.\n")
  cat("Consider using AIC or a less aggressive penalty.\n")
}
```

# Visualization: Original Effect Power Analysis

```{r power_vis_original}
#| fig-width: 12
#| fig-height: 8

# Order by AUC for visualization
plot_data <- power_results %>%
  arrange(desc(auc)) %>%
  mutate(test_label = factor(test_label, levels = test_label))

# Create the power analysis plot
p <- ggplot(plot_data, aes(x = test_label, y = sample_size, fill = auc)) +
  geom_col(color = "black", width = 0.7) +
  
  # Add sample size labels
  geom_text(
    aes(label = scales::comma(sample_size)), 
    vjust = 1.5, 
    color = "white",
    size = 5,
    fontface = "bold"
  ) +
  
  # Log scale for y-axis
  scale_y_log10(
    name = "n (sample size)",
    labels = scales::comma,
    breaks = c(10, 25, 50, 100, 250, 500, 1000, 2500, 5000, 10000, 25000, 50000)
  ) +
  
  # Color gradient
  scale_fill_gradientn(
    colors = c("#FFEBE5", "#FFBEB2", "#FA8D76", "#F46C58", "#E44144", "#C73635", "#9F3632"),
    name = "AUC",
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    guide = guide_colorbar(
      barwidth = 1,
      barheight = 15,
      title.position = "top",
      title.hjust = 0.5
    )
  ) +
  
  # Expand y-axis for labels
  coord_cartesian(clip = "off", 
                  ylim = c(min(plot_data$sample_size) * 0.7, 
                          max(plot_data$sample_size) * 2)) +
  
  # Labels and theme
  labs(
    title = "Test w/ 80% Power",
    subtitle = "Sample size needed (α = 0.05, power = 80%, two-tailed = TRUE) - BIC-Optimized Model",
    x = "",
    y = "n (sample size)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12, face = "bold", vjust = 0.5),
    axis.text.y = element_text(size = 12),
    axis.title.y = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_line(color = "gray90"),
    plot.background = element_rect(fill = "white", color = NA)
  )

print(p)

# Save plot
ggsave("multivariate_power_analysis_bic_optimized.png", p, width = 12, height = 8, dpi = 300)

# Print summary comparison
cat("\n=== POWER ANALYSIS SUMMARY ===\n")
cat("Tests included in BIC-optimal model:", nrow(power_results), "out of", length(test_codes_all), "\n")
cat("Median sample size required:", median(power_results$sample_size), "\n")
cat("Range of sample sizes:", min(power_results$sample_size), "-", max(power_results$sample_size), "\n")
cat("Tests with best power (smallest n):\n")
print(head(power_results %>% dplyr::select(test_name, sample_size, auc), 3))
```

# Full Model Power Analysis

```{r full_model_power}
# Get linear predictors from the BIC-optimized model (already calculated)
linear_predictors <- predict(glm_optimal, type = "link")

# Calculate variance of linear predictors
lp_variance <- var(linear_predictors, na.rm = TRUE)

# Calculate mean linear predictors for positive and negative outcomes
mean_lp_positive <- mean(linear_predictors[data_complete$CDPOS_W240 == 1], na.rm = TRUE)
mean_lp_negative <- mean(linear_predictors[data_complete$CDPOS_W240 == 0], na.rm = TRUE)

# Effect size: difference in mean linear predictors
effect_size <- mean_lp_positive - mean_lp_negative

# Calculate AUC
auc_obj <- roc(data_complete$CDPOS_W240, linear_predictors)
auc_value <- auc(auc_obj)

# Display the values
cat("Linear Predictor Statistics:\n")
cat("- Variance of linear predictor:", round(lp_variance, 4), "\n")
cat("- Mean LP (positive outcomes):", round(mean_lp_positive, 4), "\n")
cat("- Mean LP (negative outcomes):", round(mean_lp_negative, 4), "\n")
cat("- Effect size (difference):", round(effect_size, 4), "\n")
cat("- Model AUC:", round(auc_value, 3), "\n\n")

# Power calculation
z_alpha <- qnorm(1 - 0.05/2)  # Two-tailed test, alpha = 0.05
z_beta <- qnorm(0.80)          # 80% power

# Calculate sample size
sample_size <- ceiling(2 * lp_variance * (z_alpha + z_beta)^2 / effect_size^2)

cat("Power Calculation Components:\n")
cat("- Z-alpha (two-tailed):", round(z_alpha, 3), "\n")
cat("- Z-beta (80% power):", round(z_beta, 3), "\n")
cat("- (Z-alpha + Z-beta)^2:", round((z_alpha + z_beta)^2, 3), "\n\n")

cat("SAMPLE SIZE REQUIRED:", sample_size, "\n\n")

# Additional statistics
outcome_rate <- mean(data_complete$CDPOS_W240)
expected_events <- round(sample_size * outcome_rate)
events_per_param <- round(expected_events / 10, 1)  # 9 predictors + intercept = 10

cat("Expected outcomes with this sample size:\n")
cat("- Total participants:", sample_size, "\n")
cat("- Expected positive outcomes:", expected_events, "\n")
cat("- Events per parameter:", events_per_param, "\n\n")
```

# Visualization: Full Model Power

```{r full_model_power_vis}
# Create visualization
viz_data <- data.frame(
  Model = "BIC-Optimized Model\n(9 terms)",
  Sample_Size = sample_size,
  AUC = round(auc_value, 3),
  Effect_Size = round(effect_size, 3),
  Variance = round(lp_variance, 4)
)

# Simple bar plot
p_power <- ggplot(viz_data, aes(x = Model, y = Sample_Size)) +
  geom_col(fill = "#2E86AB", color = "black", width = 0.5) +
  
  # Sample size label on top
  geom_text(aes(label = Sample_Size), 
            vjust = -0.5, 
            size = 10, 
            fontface = "bold") +
  
  # Model statistics below
  annotate("text", x = 1, y = 5, 
           label = paste0("AUC = ", viz_data$AUC, 
                         "\nEffect Size = ", viz_data$Effect_Size,
                         "\nVariance = ", viz_data$Variance),
           size = 6, fontface = "bold") +
  
  # Y-axis scale
  scale_y_continuous(
    name = "Sample Size Required",
    limits = c(0, sample_size * 1.3),
    expand = c(0, 0)
  ) +
  
  # Labels and theme
  labs(
    title = "Power Analysis: BIC-Optimized Model",
    subtitle = "Sample size for 80% power (α = 0.05, two-tailed test)",
    x = ""
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 16, hjust = 0.5),
    axis.title.y = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 14),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    plot.margin = margin(20, 20, 20, 20)
  )

print(p_power)

# Save the plot
ggsave("bic_model_power_analysis_final.png", p_power, width = 8, height = 8, dpi = 300)

# Verification
cat("\nVERIFICATION:\n")
cat("Sample size = 2 × variance × (z_alpha + z_beta)² / effect_size²\n")
cat("Sample size = 2 ×", round(lp_variance, 4), "×", round((z_alpha + z_beta)^2, 3), "/", round(effect_size^2, 4), "\n")
cat("Sample size =", sample_size, "\n")
```

# Model Comparison Summary

```{r model_comparison}
# Create comparison between full and optimal models
model_comparison <- data.frame(
  Model = c("Full Model (All Interactions)", "AIC-Optimized Model"),
  Parameters = c(length(coef(glm_full)), length(coef(glm_optimal))),
  AIC = c(AIC(glm_full), AIC(glm_optimal)),
  BIC = c(BIC(glm_full), BIC(glm_optimal)),
  Deviance = c(deviance(glm_full), deviance(glm_optimal))
) %>%
  mutate(across(where(is.numeric), ~round(., 2)))

cat("\n\nModel Comparison:\n")
print(model_comparison)

# Summary table for power analysis
summary_table <- power_results %>%
  dplyr::select(test_name, auc, sample_size, variance, mean_delta) %>%
  mutate(
    auc = round(auc, 3),
    variance = round(variance, 4),
    mean_delta = round(mean_delta, 4)
  ) %>%
  arrange(sample_size)

kable(summary_table, 
      caption = "Power Analysis Summary - AIC-Optimized Model",
      col.names = c("Test", "AUC", "Sample Size", "Variance", "Mean Delta"),
      align = "lcccc") %>%
  kable_styling(full_width = FALSE, position = "center")
```

# Model Diagnostics and Retained Predictors

```{r model_diagnostics}
cat("\n\nOptimal Model Diagnostics:\n")
cat("==========================\n")
cat("Total observations:", nrow(data_complete), "\n")
cat("Positive outcomes:", sum(data_complete$CDPOS_W240), "\n")
cat("Outcome rate:", round(mean(data_complete$CDPOS_W240), 3), "\n")
cat("Optimal model parameters:", length(coef(glm_optimal)), "\n")
cat("AIC:", round(AIC(glm_optimal), 2), "\n")
cat("BIC:", round(BIC(glm_optimal), 2), "\n")

# Show which predictors were retained
cat("\n\nPredictors retained in optimal model:\n")
optimal_predictors <- names(coef(glm_optimal))[-1]
main_effects <- optimal_predictors[!grepl(":", optimal_predictors)]
interactions <- optimal_predictors[grepl(":", optimal_predictors)]

cat("\nMain Effects (", length(main_effects), "):\n", sep = "")
for(effect in main_effects) {
  cat("  -", effect, "\n")
}

if(length(interactions) > 0) {
  cat("\nInteractions (", length(interactions), "):\n", sep = "")
  for(int in head(interactions, 10)) {  # Show first 10
    cat("  -", int, "\n")
  }
  if(length(interactions) > 10) {
    cat("  ... and", length(interactions) - 10, "more interactions\n")
  }
}
```

# Power Analysis: 20% Effect Size Reduction

```{r power_analysis_20pct}
# Store original results for comparison
power_results_original <- power_results %>%
  mutate(
    abs_mean_delta = abs(mean_delta),
    sample_size_original = map2_dbl(variance, abs_mean_delta, 
                           ~calculate_sample_size(.x, .y, alpha = 0.05, beta = 0.20))
  )

# Calculate sample sizes for 20% REDUCTION (effect size × 1/5)
power_results <- power_results_original %>%
  mutate(
    # Reduce effect size to 1/5 of original
    abs_mean_delta_20pct = abs_mean_delta * (1/5),
    # Calculate new sample size with reduced effect
    sample_size = map2_dbl(variance, abs_mean_delta_20pct, 
                           ~calculate_sample_size(.x, .y, alpha = 0.05, beta = 0.20)),
    # Create label
    test_label = paste0(test_name, "\n(AUC: ", sprintf("%.2f", auc), ")")
  ) %>%
  filter(!is.na(sample_size) & sample_size > 0 & sample_size < 1e6) %>%
  arrange(sample_size)

cat("\nPower Analysis Results - 20% CDR Progression Reduction (BIC-Optimized Model):\n")
cat("================================================================================\n")
cat("Effect sizes reduced to 1/5 of observed to represent 20% treatment effect\n\n")

# Create comparison table
comparison_table <- power_results %>% 
  dplyr::select(test_name, auc, variance, abs_mean_delta, abs_mean_delta_20pct, 
                sample_size_original, sample_size) %>%
  mutate(
    across(c(auc, variance, abs_mean_delta, abs_mean_delta_20pct), ~round(., 3)),
    fold_increase = round(sample_size / sample_size_original, 1)
  ) %>%
  rename(
    "Test" = test_name,
    "AUC" = auc,
    "Variance" = variance,
    "Original Δ" = abs_mean_delta,
    "20% Δ" = abs_mean_delta_20pct,
    "n (Original)" = sample_size_original,
    "n (20% Effect)" = sample_size,
    "Fold Increase" = fold_increase
  )

print(comparison_table)

# Check if we have reasonable sample sizes
if(any(power_results$sample_size > 50000)) {
  cat("\nNOTE: Sample sizes >50,000 indicate very small effect sizes after 20% reduction.\n")
  cat("These tests may not be practical for detecting small treatment effects.\n")
}
```

# Visualization: 20% Effect Size Power Analysis

```{r power_vis_20pct}
#| fig-width: 12
#| fig-height: 8

# Order by AUC for visualization
plot_data <- power_results %>%
  arrange(desc(auc)) %>%
  mutate(test_label = factor(test_label, levels = test_label))

# Create the power analysis plot
p <- ggplot(plot_data, aes(x = test_label, y = sample_size, fill = auc)) +
  geom_col(color = "black", width = 0.7) +
  
  # Add sample size labels
  geom_text(
    aes(label = scales::comma(sample_size)), 
    vjust = 1.5, 
    color = "white",
    size = 5,
    fontface = "bold"
  ) +
  
  # Log scale for y-axis
  scale_y_log10(
    name = "n (sample size)",
    labels = scales::comma,
    breaks = c(100, 250, 500, 1000, 2500, 5000, 10000, 25000, 50000, 100000)
  ) +
  
  # Color gradient
  scale_fill_gradientn(
    colors = c("#FFEBE5", "#FFBEB2", "#FA8D76", "#F46C58", "#E44144", "#C73635", "#9F3632"),
    name = "AUC",
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    guide = guide_colorbar(
      barwidth = 1,
      barheight = 15,
      title.position = "top",
      title.hjust = 0.5
    )
  ) +
  
  # Expand y-axis for labels
  coord_cartesian(clip = "off", 
                  ylim = c(min(plot_data$sample_size) * 0.7, 
                          max(plot_data$sample_size) * 2)) +
  
  # Labels and theme
  labs(
    title = "Sample Size for 80% Power to Detect 20% CDR Progression Reduction",
    subtitle = "Effect sizes = 1/5 of observed (α = 0.05, power = 80%, two-tailed) - BIC-Optimized Model",
    x = "",
    y = "n (sample size)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12, face = "bold", vjust = 0.5),
    axis.text.y = element_text(size = 12),
    axis.title.y = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 13, hjust = 0.5),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_line(color = "gray90"),
    plot.background = element_rect(fill = "white", color = NA)
  )

print(p)

# Save plot
ggsave("power_analysis_20pct_reduction_bic_optimized.png", p, width = 12, height = 8, dpi = 300)

# Print summary
cat("\n=== POWER ANALYSIS SUMMARY - 20% EFFECT ===\n")
cat("Tests included in BIC-optimal model:", nrow(power_results), "out of", length(test_codes_all), "\n")
cat("Median sample size required:", median(power_results$sample_size), "\n")
cat("Range of sample sizes:", min(power_results$sample_size), "-", max(power_results$sample_size), "\n")
cat("\nTests with best power for 20% effect (smallest n):\n")
print(head(power_results %>% dplyr::select(test_name, sample_size, auc, abs_mean_delta_20pct), 3))
```

# CFISP Effect Size Calibration Analysis

```{r cfisp_calibration_setup}
# Function to find the multiplier that gives a target sample size
find_effect_multiplier <- function(variance, original_delta, target_n, alpha = 0.05, beta = 0.20) {
  z_alpha <- qnorm(1 - alpha / 2)
  z_beta <- qnorm(1 - beta)
  
  # From n = (2 * variance * (z_alpha + z_beta)^2) / delta^2
  # Solving for delta: delta = sqrt((2 * variance * (z_alpha + z_beta)^2) / n)
  required_delta <- sqrt((2 * variance * (z_alpha + z_beta)^2) / target_n)
  
  # The multiplier is: required_delta / original_delta
  multiplier <- required_delta / abs(original_delta)
  
  return(multiplier)
}

# Find CFISP in your original results
cfisp_data <- power_results_original %>%
  filter(test_name == "CFISP")

if(nrow(cfisp_data) > 0) {
  # Extract CFISP parameters
  cfisp_variance <- cfisp_data$variance[1]
  cfisp_original_delta <- cfisp_data$abs_mean_delta[1]
  
  cat("CFISP Original Values:\n")
  cat("=====================\n")
  cat("Original delta:", round(cfisp_original_delta, 4), "\n")
  cat("Variance:", round(cfisp_variance, 4), "\n")
  cat("Current n (with 1/5 multiplier):", 635, "\n")
  cat("Target n:", 396, "\n\n")
  
  # Find the multiplier for n = 396
  target_n <- 396
  custom_multiplier <- find_effect_multiplier(cfisp_variance, cfisp_original_delta, target_n)
  
  cat("SOLUTION:\n")
  cat("=========\n")
  cat("Required multiplier:", round(custom_multiplier, 4), "\n")
  cat("This represents a", round((1 - custom_multiplier) * 100, 1), "% reduction in effect size\n\n")
  
  # Verify the calculation
  new_delta <- cfisp_original_delta * custom_multiplier
  verify_n <- calculate_sample_size(cfisp_variance, new_delta)
  cat("Verification - CFISP n with new multiplier:", verify_n, "\n\n")
} else {
  stop("CFISP not found in results!")
}
```

```{r apply_custom_multiplier}
# Now apply this custom multiplier to ALL tests
cat("Applying multiplier", round(custom_multiplier, 4), "to all tests:\n")
cat("==================================================\n\n")

# Recalculate sample sizes with the custom multiplier
power_results <- power_results_original %>%
  mutate(
    # Apply the CFISP-calibrated multiplier to all tests
    abs_mean_delta_custom = abs_mean_delta * custom_multiplier,
    # Calculate new sample size
    sample_size = map2_dbl(variance, abs_mean_delta_custom, 
                           ~calculate_sample_size(.x, .y, alpha = 0.05, beta = 0.20)),
    # Create label
    test_label = paste0(test_name, "\n(AUC: ", sprintf("%.2f", auc), ")")
  ) %>%
  filter(!is.na(sample_size) & sample_size > 0 & sample_size < 1e6) %>%
  arrange(sample_size)

# Create comparison table
comparison_table <- power_results %>% 
  dplyr::select(test_name, auc, variance, abs_mean_delta, abs_mean_delta_custom, sample_size) %>%
  mutate(
    across(c(auc, variance, abs_mean_delta, abs_mean_delta_custom), ~round(., 4)),
    effect_reduction = paste0(round((1 - custom_multiplier) * 100, 1), "%")
  ) %>%
  rename(
    "Test" = test_name,
    "AUC" = auc,
    "Variance" = variance,
    "Original Δ" = abs_mean_delta,
    "New Δ" = abs_mean_delta_custom,
    "n" = sample_size,
    "Effect Reduction" = effect_reduction
  )

print(comparison_table)

# Specifically highlight CFISP to confirm
cat("\n\nConfirming CFISP sample size:\n")
cfisp_final <- comparison_table %>% filter(Test == "CFISP")
print(cfisp_final)
```

# Visualization: CFISP-Calibrated Power Analysis

```{r cfisp_calibrated_vis}
#| fig-width: 12
#| fig-height: 8

# Order by AUC for visualization
plot_data <- power_results %>%
  arrange(desc(auc)) %>%
  mutate(test_label = factor(test_label, levels = test_label))

# Create the power analysis plot with custom multiplier
p <- ggplot(plot_data, aes(x = test_label, y = sample_size, fill = auc)) +
  geom_col(color = "black", width = 0.7) +
  
  # Add sample size labels
  geom_text(
    aes(label = scales::comma(sample_size)), 
    vjust = 1.5, 
    color = "white",
    size = 5,
    fontface = "bold"
  ) +
  
  # Log scale for y-axis
  scale_y_log10(
    name = "n (sample size)",
    labels = scales::comma,
    breaks = c(100, 250, 500, 1000, 2500, 5000, 10000, 25000, 50000, 100000, 250000)
  ) +
  
  # Color gradient
  scale_fill_gradientn(
    colors = c("#FFEBE5", "#FFBEB2", "#FA8D76", "#F46C58", "#E44144", "#C73635", "#9F3632"),
    name = "AUC",
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    guide = guide_colorbar(
      barwidth = 1,
      barheight = 15,
      title.position = "top",
      title.hjust = 0.5
    )
  ) +
  
  # Expand y-axis for labels
  coord_cartesian(clip = "off", 
                  ylim = c(min(plot_data$sample_size) * 0.7, 
                          max(plot_data$sample_size) * 2)) +
  
  # Labels and theme
  labs(
    title = paste0("Sample Size for 80% Power"),
    subtitle = paste0("To Detect ", round((custom_multiplier) * 100, 1), 
                     "% CDR Progression Reduction"),
    x = "",
    y = "n (sample size)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12, face = "bold", vjust = 0.5),
    axis.text.y = element_text(size = 12),
    axis.title.y = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 13, hjust = 0.5),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_line(color = "gray90"),
    plot.background = element_rect(fill = "white", color = NA)
  )

print(p)

# Save plot
ggsave("power_analysis_cfisp_calibrated.png", p, width = 12, height = 8, dpi = 300)

# Print summary
cat("\n=== POWER ANALYSIS SUMMARY - CFISP CALIBRATED ===\n")
cat("Multiplier used:", round(custom_multiplier, 4), "\n")
cat("Effect reduction:", round((1 - custom_multiplier) * 100, 1), "%\n")
cat("CFISP sample size:", power_results %>% filter(test_name == "CFISP") %>% pull(sample_size), "\n")
cat("Median sample size across all tests:", median(power_results$sample_size), "\n")
cat("Range of sample sizes:", min(power_results$sample_size), "-", max(power_results$sample_size), "\n")
```

# Full Model Custom Effect Size Analysis

```{r full_model_custom}
# Get linear predictors from the BIC-optimized model
linear_predictors <- predict(glm_optimal, type = "link")

# Calculate variance of linear predictors
lp_variance <- var(linear_predictors, na.rm = TRUE)

# Calculate mean linear predictors for positive and negative outcomes
mean_lp_positive <- mean(linear_predictors[data_complete$CDPOS_W240 == 1], na.rm = TRUE)
mean_lp_negative <- mean(linear_predictors[data_complete$CDPOS_W240 == 0], na.rm = TRUE)

# Original effect size
effect_size_original <- mean_lp_positive - mean_lp_negative

# CUSTOM effect size (using the multiplier that makes CFISP n=396)
effect_size_custom <- effect_size_original * custom_multiplier

# Calculate AUC
auc_obj <- roc(data_complete$CDPOS_W240, linear_predictors)
auc_value <- auc(auc_obj)

# Power calculation with custom effect size
z_alpha <- qnorm(1 - 0.05/2)  # Two-tailed test, alpha = 0.05
z_beta <- qnorm(0.80)          # 80% power

# Calculate sample size with custom multiplier
sample_size_custom <- ceiling(2 * lp_variance * (z_alpha + z_beta)^2 / effect_size_custom^2)

# Create single bar visualization
viz_data <- data.frame(
  Model = paste0(round((custom_multiplier) * 100, 1), "% Effect"),
  Sample_Size = sample_size_custom,
  Effect_Size = round(effect_size_custom, 3),
  Variance = round(lp_variance, 4),
  AUC = round(auc_value, 3)
)

# Single bar plot
p_power_single <- ggplot(viz_data, aes(x = Model, y = Sample_Size)) +
  geom_col(fill = "#E74C3C", color = "black", width = 0.5) +
  
  # Sample size label on top
  geom_text(aes(label = scales::comma(Sample_Size)), 
            vjust = -0.5, 
            size = 10, 
            fontface = "bold") +
  
  # Statistics in the middle of the bar
  annotate("text", x = 1, y = sample_size_custom/2, 
           label = paste0("AUC = ", viz_data$AUC, "\n",
                         "Effect = ", viz_data$Effect_Size, "\n",
                         "Variance = ", viz_data$Variance),
           size = 6, 
           fontface = "bold",
           color = "white") +
  
  # Y-axis scale
  scale_y_continuous(
    name = "Sample Size Required",
    labels = scales::comma,
    limits = c(0, sample_size_custom * 1.3),
    expand = c(0, 0)
  ) +
  
  # Labels and theme
  labs(
    title = "Power Analysis: Full BIC-Optimized Model",
    subtitle = "Sample size for 80% power (α = 0.05, two-tailed test)",
    x = ""
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 16, hjust = 0.5),
    axis.title.y = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 14),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    plot.margin = margin(20, 20, 20, 20)
  )

print(p_power_single)

# Save the plot
ggsave("bic_model_power_analysis_custom_single.png", p_power_single, width = 9, height = 5, dpi = 300)

# Print summary
cat("\nPower Analysis Summary:\n")
cat("======================\n")
cat("Sample size required:", sample_size_custom, "\n")
cat("Model AUC:", round(auc_value, 3), "\n")
cat("Effect size:", round(effect_size_custom, 3), "\n")
cat("Variance:", round(lp_variance, 4), "\n")
```

# Model Coefficients and Odds Ratios

```{r model_coefficients}
# Calculate odds ratios and confidence intervals
# Extract coefficients and confidence intervals
coef_summary <- summary(glm_optimal)$coefficients

# Calculate odds ratios
odds_ratios <- exp(coef(glm_optimal))

# Calculate 95% confidence intervals for odds ratios
conf_int <- exp(confint(glm_optimal))

# Create a clean summary table
or_table <- data.frame(
  Variable = names(coef(glm_optimal)),
  Coefficient = coef(glm_optimal),
  OR = odds_ratios,
  CI_Lower = conf_int[,1],
  CI_Upper = conf_int[,2],
  P_Value = coef_summary[,4]
) %>%
  mutate(
    # Format the output
    OR_formatted = round(OR, 3),
    CI_formatted = paste0("(", round(CI_Lower, 3), ", ", round(CI_Upper, 3), ")"),
    P_formatted = formatp(P_Value),
    # Add significance stars
    Significance = case_when(
      P_Value < 0.001 ~ "***",
      P_Value < 0.01 ~ "**",
      P_Value < 0.05 ~ "*",
      P_Value < 0.1 ~ ".",
      TRUE ~ ""
    )
  )

# Display the results
cat("\n=== ODDS RATIOS FOR BIC-OPTIMIZED MODEL ===\n")
print(or_table %>% 
      dplyr::select(Variable, OR_formatted, CI_formatted, P_formatted, Significance) %>%
      rename("Odds Ratio" = OR_formatted,
             "95% CI" = CI_formatted,
             "p-value" = P_formatted,
             "Sig." = Significance))

# For a cleaner display of just the key information
cat("\n\n=== SIMPLIFIED OUTPUT ===\n")
or_table %>%
  filter(Variable != "(Intercept)") %>%
  mutate(
    Effect = paste0(round(OR, 3), " ", CI_formatted, " ", Significance)
  ) %>%
  dplyr::select(Variable, Effect) %>%
  print()
```

# Overall Model Performance Metrics

```{r model_performance}
# Calculate overall model odds ratio (Nagelkerke R²) and performance metrics
# Get model predictions
predictions <- predict(glm_optimal, type = "response")

# Calculate AUC with confidence interval
roc_obj <- roc(data_complete$CDPOS_W240, predictions)
auc_ci <- ci.auc(roc_obj, conf.level = 0.95)

# Calculate Nagelkerke R²
null_model <- glm(CDPOS_W240 ~ 1, data = data_complete, family = binomial)
n <- nrow(data_complete)
R2_nagelkerke <- (1 - exp((logLik(glm_optimal) - logLik(null_model)) * (-2/n))) / 
                 (1 - exp(logLik(null_model) * (-2/n)))

# Calculate overall discrimination slope (mean predicted probability difference)
mean_pred_pos <- mean(predictions[data_complete$CDPOS_W240 == 1])
mean_pred_neg <- mean(predictions[data_complete$CDPOS_W240 == 0])
discrimination_slope <- mean_pred_pos - mean_pred_neg

# Display results
cat("\n=== OVERALL MODEL PERFORMANCE ===\n")
cat("AUC: ", round(auc_ci[2], 3), " (95% CI: ", round(auc_ci[1], 3), "-", round(auc_ci[3], 3), ")\n", sep = "")
cat("Nagelkerke R²: ", round(R2_nagelkerke, 3), "\n", sep = "")
cat("Discrimination slope: ", round(discrimination_slope, 3), "\n", sep = "")
cat("Number of observations: ", n, "\n", sep = "")
cat("Number of events: ", sum(data_complete$CDPOS_W240), "\n", sep = "")
```

# Final Model Comparison Visualization

```{r final_comparison}
# Create data frame with model information
models_data <- data.frame(
  Model = factor(c("Uni-test:\nCFITOTAL", 
                   "Multi-test:\nCFISP", 
                   "Multi-test Full Model:\nCFISP, ADLPQPT, LMIIA,\nBaselines, C3"),
                 levels = c("Uni-test:\nCFITOTAL", 
                           "Multi-test:\nCFISP", 
                           "Multi-test Full Model:\nCFISP, ADLPQPT, LMIIA,\nBaselines, C3")),
  Sample_Size = c(85, 635, 302),
  AUC = c(0.74, 0.74, 0.835),
  OR_Lower = c(1.44, 0.687, 0.778),
  OR_Upper = c(3.14, 1.3, 0.892)
)

# Create the bar plot
p <- ggplot(models_data, aes(x = Model, y = Sample_Size)) +
  geom_bar(stat = "identity", fill = c("#3498db", "#e74c3c", "#e74c3c"), width = 0.7) +
  
  # Add sample size labels on top of bars
  geom_text(aes(label = Sample_Size), 
            vjust = -0.5, 
            size = 6, 
            fontface = "bold") +
  
  # Add AUC and OR information inside bars
  geom_text(aes(y = Sample_Size/2, 
                label = paste0("AUC = ", AUC, "\n",
                              "OR = [", OR_Lower, ", ", OR_Upper, "]")),
            color = "white", 
            size = 4, 
            fontface = "bold") +
  
  # Customize the plot
  labs(title = "Power Analysis: Model Comparison",
       subtitle = "Sample size for 80% power (α = 0.05, two-tailed test)",
       x = "",
       y = "Sample Size Required") +
  
  # Set y-axis limits
  scale_y_continuous(limits = c(0, 700), 
                     breaks = seq(0, 700, 100)) +
  
  # Apply theme
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5, color = "gray40"),
    axis.text.x = element_text(size = 11, face = "bold"),
    axis.text.y = element_text(size = 11),
    axis.title.y = element_text(size = 12, face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

# Display the plot
print(p)

# Save the plot (optional)
ggsave("power_analysis_comparison.png", 
       plot = p, 
       width = 7, 
       height = 7, 
       dpi = 300, 
       bg = "white")

# Print summary table
cat("\n=== Model Comparison Summary ===\n")
print(models_data, row.names = FALSE)
```
