---
title: "5 Predictive Models & Full ROC Curve"
author: "Ian Sequeira"
format: html
---

# Load Libraries and Helper Functions

```{r setup}
#| message: false
#| warning: false

library(here)
source(here("helpers", "01-load-libraries.R"))
source(here("helpers", "02-functions.R"))
```

# Custom Functions

```{r ROC functions + AUC Calc}
#| message: false
#| warning: false

# Function to calculate ROC
calculate_roc <- function(y_true, y_pred_prob) {
  # Remove any NA values
  complete_idx <- complete.cases(y_true, y_pred_prob)
  y_true <- y_true[complete_idx]
  y_pred_prob <- y_pred_prob[complete_idx]
  
  # Check if we have valid data
  if(length(y_true) == 0 || length(unique(y_true)) < 2) {
    warning("Insufficient data for ROC calculation")
    return(list(
      sensitivity = c(0, 1),
      specificity = c(1, 0),
      fpr = c(0, 1),
      auc = 0.5
    ))
  }
  
  cutpoints <- seq(0, 1, length.out = 1000)
  sensitivity <- specificity <- rep(NA, length(cutpoints))
  
  for(i in 1:length(cutpoints)) {
    pred_class <- ifelse(y_pred_prob >= cutpoints[i], 1, 0)
    
    TP <- sum(y_true == 1 & pred_class == 1, na.rm = TRUE)
    TN <- sum(y_true == 0 & pred_class == 0, na.rm = TRUE)
    FP <- sum(y_true == 0 & pred_class == 1, na.rm = TRUE)
    FN <- sum(y_true == 1 & pred_class == 0, na.rm = TRUE)
    
    sensitivity[i] <- if((TP + FN) > 0) TP / (TP + FN) else 0
    specificity[i] <- if((TN + FP) > 0) TN / (TN + FP) else 0
  }
  
  # Calculate AUC using trapezoidal rule
  auc <- sum(diff(1 - specificity) * (sensitivity[-1] + sensitivity[-length(sensitivity)]) / 2)
  
  return(list(
    sensitivity = sensitivity,
    specificity = specificity,
    fpr = 1 - specificity, 
    auc = abs(auc)
  ))
}
```

# Data Preparation & LEARN Effect Adjustment

```{r data_prep}
#| message: false
#| warning: false

# Load data files
data_wk48 <- read_csv("data/data_wk48.csv")
learn_data <- read_csv("data/data_learn&base_wk48.csv")

# Calculate LEARN effects
learn_effects <- learn_data %>%
  summarise(
    across(starts_with("z_delta_") & ends_with("_W48"), 
           ~mean(.x, na.rm = TRUE), 
           .names = "learn_effect_{.col}")
  ) %>%
  rename_all(~str_replace(., "z_delta_", ""))

# Create adjusted deltas and baselines
data_wk48_adjusted <- data_wk48 %>%
  mutate(
    # Baselines
    baseline_ADLPQPT = ADLPQPT_W48,
    baseline_ADLPQSP = ADLPQSP_W48,
    baseline_CFIPT = CFIPT_W48,
    baseline_CFISP = CFISP_W48,
    baseline_DIGIT = DIGIT_W48,
    baseline_FCSRT96 = FCSRT96_W48,
    baseline_LMIIA = LMIIA_W48,
    baseline_MMSE = MMSE_W48,
    # Adjusted deltas
    adjusted_delta_ADLPQPT_W48 = z_delta_ADLPQPT_W48 - learn_effects$learn_effect_ADLPQPT_W48,
    adjusted_delta_ADLPQSP_W48 = z_delta_ADLPQSP_W48 - learn_effects$learn_effect_ADLPQSP_W48,
    adjusted_delta_CFIPT_W48 = z_delta_CFIPT_W48 - learn_effects$learn_effect_CFIPT_W48,
    adjusted_delta_CFISP_W48 = z_delta_CFISP_W48 - learn_effects$learn_effect_CFISP_W48,
    adjusted_delta_DIGIT_W48 = z_delta_DIGIT_W48 - learn_effects$learn_effect_DIGIT_W48,
    adjusted_delta_FCSRT96_W48 = z_delta_FCSRT96_W48 - learn_effects$learn_effect_FCSRT96_W48,
    adjusted_delta_LMIIA_W48 = z_delta_LMIIA_W48 - learn_effects$learn_effect_LMIIA_W48,
    adjusted_delta_MMSE_W48 = z_delta_MMSE_W48 - learn_effects$learn_effect_MMSE_W48
  )

# Remove rows with missing outcomes
data_complete <- data_wk48_adjusted %>% 
  filter(!is.na(CDPOS_W240))

cat("Data dimensions after processing:", nrow(data_complete), "x", ncol(data_complete), "\n")
```

```{r variable_setup}
#| message: false
#| warning: false

# Get variable lists
adjusted_vars <- grep("^adjusted_delta_.*_W48$", names(data_complete), value = TRUE)
if("z_delta_C3_W48" %in% names(data_complete)) {
  adjusted_vars <- c(adjusted_vars, "z_delta_C3_W48")
}
baseline_vars <- grep("^baseline_", names(data_complete), value = TRUE)

cat("Predictors available:\n")
cat("Adjusted deltas:", length(adjusted_vars), "\n")
cat("Baseline covariates:", length(baseline_vars), "\n\n")
```

# Model 1: GLM Main Effects Only

```{r model_glm_main}
#| message: false
#| warning: false

cat("Fitting Model 1: GLM with main effects only...\n")
formula_main <- as.formula(paste("CDPOS_W240 ~", 
                                paste(c(adjusted_vars, baseline_vars), 
                                      collapse = " + ")))
glm_main <- glm(formula_main, data = data_complete, family = binomial(link = "logit"))
```

# Model 2: Best Subsets Selection

```{r model_best_subsets}
#| message: false
#| warning: false

cat("Fitting Model 2: Best Subsets...\n")
all_vars <- c(adjusted_vars, baseline_vars)
model_data_bs <- na.omit(data_complete[, c("CDPOS_W240", all_vars)])
X_bs <- model_data_bs[, -1]
y_bs <- model_data_bs$CDPOS_W240

regfit <- regsubsets(X_bs, y_bs, nvmax = min(ncol(X_bs), 20), method = "exhaustive")
best_bic <- which.min(summary(regfit)$bic)
selected_vars <- names(coef(regfit, best_bic))[-1]

if(length(selected_vars) > 0) {
  formula_bs <- as.formula(paste("CDPOS_W240 ~", paste(selected_vars, collapse = " + ")))
  glm_bs <- glm(formula_bs, data = model_data_bs, family = binomial)
} else {
  glm_bs <- glm(CDPOS_W240 ~ 1, data = model_data_bs, family = binomial)
}
```

# Model 3: Ridge Regression

```{r model_ridge}
#| message: false
#| warning: false

cat("Fitting Model 3: Ridge Regression...\n")
X_ridge <- as.matrix(data_complete[, all_vars])
y_ridge <- data_complete$CDPOS_W240
set.seed(123)
cv_ridge <- cv.glmnet(X_ridge, y_ridge, alpha = 0, family = "binomial", nfolds = 5)
```

# Model 4: CART (Decision Tree)

```{r model_cart}
#| message: false
#| warning: false

cat("Fitting Model 4: CART...\n")
set.seed(123)
cart_full <- rpart(CDPOS_W240 ~ ., 
                   data = data_complete[, c("CDPOS_W240", all_vars)], 
                   method = "class", cp = 0, minsplit = 5)
cp_table <- cart_full$cptable
best_row <- which.min(cp_table[, "xerror"])
cp_1se <- cp_table[cp_table[, "xerror"] <= cp_table[best_row, "xerror"] + 
                   cp_table[best_row, "xstd"], "CP"][1]
cart_pruned <- prune(cart_full, cp = cp_1se)
```

# Model 5: GLM Full Interactions

```{r model_glm_full}
#| message: false
#| warning: false

cat("Fitting Model 5: GLM with all 2-way interactions...\n")
formula_full <- as.formula(paste("CDPOS_W240 ~ (", 
                                paste(adjusted_vars, collapse = " + "),
                                ")^2 + ",
                                paste(baseline_vars, collapse = " + ")))
glm_full <- glm(formula_full, data = data_complete, family = binomial(link = "logit"))
cat("Full model parameters:", length(coef(glm_full)), "\n")
```

# Model 6: GLM Optimal (BIC-Selected)

```{r model_glm_optimal}
#| message: false
#| warning: false

cat("Fitting Model 6: GLM Optimal (BIC-selected)...\n")
cat("Running stepwise selection...\n")
glm_optimal <- stepAIC(glm_full, 
                      direction = "both",
                      k = log(nrow(data_complete)),  # BIC penalty
                      trace = FALSE)
cat("Optimal model parameters:", length(coef(glm_optimal)), "\n")
cat("Parameters removed:", length(coef(glm_full)) - length(coef(glm_optimal)), "\n\n")
```

# Model Predictions & ROC Calculations

```{r roc_calculations}
# Calculate predictions and ROCs for all models
cat("Calculating ROC curves...\n")

# 1. GLM Main Effects
pred_glm_main <- predict(glm_main, type = "response")
roc_glm_main <- calculate_roc(data_complete$CDPOS_W240, pred_glm_main)

# 2. Best Subsets
pred_bs <- predict(glm_bs, newdata = model_data_bs, type = "response")
roc_bs <- calculate_roc(model_data_bs$CDPOS_W240, pred_bs)

# 3. Ridge
pred_ridge <- predict(cv_ridge, newx = X_ridge, s = "lambda.min", type = "response")[,1]
roc_ridge <- calculate_roc(y_ridge, pred_ridge)

# 4. CART
pred_cart <- predict(cart_pruned, type = "prob")[, 2]
roc_cart <- calculate_roc(data_complete$CDPOS_W240, pred_cart)

# 5. GLM Full
pred_glm_full <- predict(glm_full, type = "response")
roc_glm_full <- calculate_roc(data_complete$CDPOS_W240, pred_glm_full)

# 6. GLM Optimal
pred_glm_optimal <- predict(glm_optimal, type = "response")
roc_glm_optimal <- calculate_roc(data_complete$CDPOS_W240, pred_glm_optimal)
```

# ROC Curve Visualization

```{r roc_plot}
#| fig-cap: "ROC Curves for Six Models - Week 48"

# Create the ROC plot
plot(roc_glm_main$fpr, roc_glm_main$sensitivity, type = "l", lwd = 2.5, col = "black",
     main = "ROC Curves - Week 48\n(All models include baseline + LEARN-adjusted deltas)",
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     xlim = c(0, 1), ylim = c(0, 1),
     cex.main = 1.4, cex.lab = 1.2, cex.axis = 1.1)

# Add other curves
lines(roc_bs$fpr, roc_bs$sensitivity, lwd = 2, col = "purple", lty = 2)
lines(roc_ridge$fpr, roc_ridge$sensitivity, lwd = 2, col = "blue", lty = 3)
lines(roc_cart$fpr, roc_cart$sensitivity, lwd = 2, col = "darkgreen", lty = 4)
lines(roc_glm_full$fpr, roc_glm_full$sensitivity, lwd = 2, col = "red", lty = 5)
lines(roc_glm_optimal$fpr, roc_glm_optimal$sensitivity, lwd = 3, col = "red")

# Add diagonal reference line
abline(a = 0, b = 1, lty = 2, col = "gray")

# Add legend with sections
legend("bottomright", 
       legend = c("Non-Cross-Validated:",
                  paste0("  GLM Main Effects (AUC = ", round(roc_glm_main$auc, 3), ")"),
                  paste0("  Best Subsets (AUC = ", round(roc_bs$auc, 3), ")"),
                  paste0("  GLM Full 2-way (AUC = ", round(roc_glm_full$auc, 3), ")"),
                  paste0("  GLM Optimal (AUC = ", round(roc_glm_optimal$auc, 3), ")"),
                  "",  # Spacer
                  "Cross-Validated:",
                  paste0("  Ridge (AUC = ", round(roc_ridge$auc, 3), ")"),
                  paste0("  CART (AUC = ", round(roc_cart$auc, 3), ")")),
       col = c("black",
               "black", "purple", "red", "red",
               "white",
               "black",
               "blue", "darkgreen"),
       lty = c(NA,
               1, 2, 5, 1,
               NA,
               NA,
               3, 4),
       lwd = c(NA,
               2.5, 2, 2, 3,
               NA,
               NA,
               2, 2),
       cex = 0.5,         # Smaller font
       bg = "white",      # Background
       bty = "o",         # Box around legend
       text.font = c(2,   # Bold header
                     1, 1, 1, 1,
                     1,
                     2,
                     1, 1))
```

# Model Performance Summary

```{r performance_summary}
# Create AUC summary table
auc_summary <- data.frame(
  Model = c("GLM Main Effects", "Best Subsets", "Ridge Regression", 
            "CART", "GLM Full 2-way", "GLM Optimal"),
  Parameters = c(length(coef(glm_main)), 
                 length(selected_vars) + 1,
                 sum(coef(cv_ridge, s = "lambda.min") != 0),
                 sum(cart_pruned$frame$var != "<leaf>"),
                 length(coef(glm_full)),
                 length(coef(glm_optimal))),
  AUC = c(roc_glm_main$auc, roc_bs$auc, roc_ridge$auc, 
          roc_cart$auc, roc_glm_full$auc, roc_glm_optimal$auc)
) %>%
  mutate(AUC = round(AUC, 3)) %>%
  arrange(desc(AUC))

cat("\n=== MODEL PERFORMANCE SUMMARY ===\n")
print(auc_summary)

# Highlight best performing model
best_model <- auc_summary$Model[1]
cat("\nBest performing model:", best_model, "with AUC =", auc_summary$AUC[1], "\n")

# Show parameter reduction for GLM models
cat("\n=== GLM MODEL COMPARISON ===\n")
cat("GLM Main Effects:", length(coef(glm_main)), "parameters\n")
cat("GLM Full 2-way:", length(coef(glm_full)), "parameters\n")
cat("GLM Optimal:", length(coef(glm_optimal)), "parameters\n")
cat("Parameter reduction:", 
    round((1 - length(coef(glm_optimal))/length(coef(glm_full))) * 100, 1), "%\n")
```

# Export High-Quality Plot

```{r export_plot}
# Save the plot as PNG
png("roc_curves_six_models_week48.png", width = 10, height = 8, units = "in", res = 300)

# Create the ROC plot
plot(roc_glm_main$fpr, roc_glm_main$sensitivity, type = "l", lwd = 2.5, col = "black",
     main = "ROC Curves - Week 48",
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     xlim = c(0, 1), ylim = c(0, 1),
     cex.main = 1.4, cex.lab = 1.2, cex.axis = 1.1)

# Add other curves
lines(roc_bs$fpr, roc_bs$sensitivity, lwd = 2, col = "brown", lty = 2)
lines(roc_ridge$fpr, roc_ridge$sensitivity, lwd = 2, col = "blue", lty = 3)
lines(roc_cart$fpr, roc_cart$sensitivity, lwd = 2, col = "darkgreen", lty = 4)
lines(roc_glm_full$fpr, roc_glm_full$sensitivity, lwd = 2, col = "orange", lty = 5)
lines(roc_glm_optimal$fpr, roc_glm_optimal$sensitivity, lwd = 5, col = "red")

# Add diagonal reference line
abline(a = 0, b = 1, lty = 2, col = "gray")

# Add legend with sections
legend("bottomright", 
       legend = c("Non-Cross-Validated:",
                  paste0("  GLM Main Effects (AUC = ", round(roc_glm_main$auc, 3), ")"),
                  paste0("  Best Subsets (AUC = ", round(roc_bs$auc, 3), ")"),
                  paste0("  GLM Full 2-way (AUC = ", round(roc_glm_full$auc, 3), ")"),
                  paste0("  GLM Optimal (AUC = ", round(roc_glm_optimal$auc, 3), ")"),
                  "",  # Spacer
                  "Cross-Validated:",
                  paste0("  Ridge (AUC = ", round(roc_ridge$auc, 3), ")"),
                  paste0("  CART (AUC = ", round(roc_cart$auc, 3), ")")),
       col = c("black",
               "black", "brown", "orange", "red",
               "white",
               "black",
               "blue", "darkgreen"),
       lty = c(NA,
               1, 2, 5, 1,
               NA,
               NA,
               3, 4),
       lwd = c(NA,
               2.5, 2, 2, 5,
               NA,
               NA,
               2, 2),
       cex = 0.9,         # Smaller font
       bg = "white",      # Background
       bty = "o",         # Box around legend
       text.font = c(2,   # Bold header
                     1, 1, 1, 1,
                     1,
                     2,
                     1, 1))
# Close the PNG device
dev.off()

cat("Plot saved as: roc_curves_six_models_week48.png\n")
```